Haul-out behavior and aerial survey detectability of seals in the Bering and Chukchi seas

### 2021-04-07 analysis/paper/paper.Rmd
INITIAL COMMENTS FROM MFC

Ice-associated seals rely on sea ice for a variety of activities, including breeding,
molting, pupping, and resting. In the Arctic, many of these activities occur in spring and
early summer (April-June) as sea ice begins to melt and retreat northward. Rapid
acceleration of climate change in Arctic ecosystems is thus of concern as the quantity and
quality of suitable habitat is forecast to decrease. In this paper, we use data collected
from satellite-linked telemetry tags deployed between 2005 and 2020 to investigate the
seasonal timing and environmental factors affecting sea ice use by seals (specifically,
bearded, ribbon, and spotted seals) in the Bering, and Chukchi seas. In addition to
providing baseline data on phenology, these data also allow us to refine availability
estimates needed to accurately estimate abundance from aerial survey counts of seals
basking on ice (i.e., to correct for the proportion of animals that are in the water while
surveys are conducted). Using generalized linear mixed pseudo-models to properly account
for temporal autocorrelation, we fit models with a variety of covariates (e.g.,
day-of-year, solar hour, age-sex class, wind speed, barometric pressure, temperature,
precipitation) to examine their ability to explain variation in hourly haul-out records.
We found evidence for strong diurnal and within season patterns in haul-out behavior, as
well as strong weather effects (particularly wind and temperature). In general, seals were
more likely to rest on ice in the middle of the day and when wind speed was low and
temperatures were higher. Haul-out propensity increased through March and April, peaking
in May and early June before declining again. The timing and frequency of haul-out events
also varied based on species and age-sex class. For ribbon and spotted seals, models with
year effects were highly supported, indicating that the timing and magnitude of haul-out
behavior varied among years. However, we did not find evidence that haul-out
timing was linked annual sea ice extent. Our analysis emphasizes the importance of accounting
for variation in haul-out behavior when interpreting the number of seals
counted in aerial surveys.

@@

For ice-associated seals, haul out is tightly linked with sea ice. Ribbon seals rest out
of the water almost exclusively on sea ice and are mostly pelagic outside the spring
pupping, breeding, and molting season [@burns1981]. While spotted and bearded seals do
rest on beaches during ice-free months, they strongly prefer sea ice as a resting platform
during the spring and early summer [@burns2009]. The very remote nature of the Bering and
Chukchi seas mean direct observation of seal behavior is limited. Bio-logging devices are,
thus, especially useful tools for collecting key information on movement and haul-out
behavior for these species. Records of time spent out of the water provide valuable data
for identifying covariates that explain variation in haul-out behavior. For instance, in
the Antarctic, Bengtson and Cameron [-@bengtson2004pb] relied on bio-logging data to
demonstrate greater haul-out propensity in juvenile crabeater seals (*Lobodon
carcinophaga*) than adults, with highest probabilities in February and at times close to
solar noon. In the Arctic, Von Duyke et al. [-@vonduyke2020ee] demonstrated seasonal
changes between diurnal and nocturnal haul-out behavior of ringed seals. Bengtson et al.
[-@bengtson2005pb] documented a higher propensity for ringed seal basking near solar noon,
as did Ver Hoef et al. [-@verhoef2014sm] in an analysis of bearded (*Erignathus
barbatus*), ribbon (*Histriophoca fasciata*), and spotted (*Phoca largha*) seals using
much larger sample sizes. Olnes et al. [-@olnes2020pb] showed that the proportion of time
bearded seals spent hauled out progressively increased through spring and summer. And, Ver
Hoef et al. [-@verhoef2014sm] found haul-out probabilities increased gradually starting in
March and peaking in May and June for bearded, ribbon, and spotted seals. However, their
primary focus was estimating seal abundance, and like a subsequent analysis by Conn et al.
[-@conn2014mee], they did not attempt to separately analyze haul-out schedules for
different age and sex classes (although initial models with sex and age as fixed effects
were insignificant at p = 0.05; [@verhoef2014sm]).

@@

Ultimately, knowledge of trends in phenology and abundance (or life history surrogates
such as survival and recruitment) will be necessary to make credible quantitative
predictions about the effects of the climate crisis on the abundance and distribution of
Arctic seal populations. Before we can construct a trend, however, we first need credible
data points. Several studies have contributed estimates of the distribution and abundance
of ice-associated seal species in the Arctic using aerial surveys (e.g.,
[@bengtson2005pb], [@conn2014mee], and [@verhoef2014sm]). Such studies are conducted over
very large areas and estimation of absolute abundance requires making inference about
numerous nuisance processes affecting the observation of seals on ice. This includes
availability (only seals basking on ice are available to be counted), detection
probability (observers or automated detection systems may miss some seals on ice), species
misclassification, and possible disturbance of seals by aircraft [@conn2014mee;
@verhoef2014sm].

@@

Haul-out behavior data from `r dplyr::n_distinct(deploy_tbl$speno)` bio-loggers deployed
on bearded, spotted, and ribbon seals was subset to include only records from 1 March to
15 July between `r min(lubridate::year(dat.sf$haulout_dt))` and
`r max(lubridate::year(dat.sf$haulout_dt))`. Bio-loggers were of the 'SPLASH' or 'SPOT'
family of tags developed by Wildlife Computers (Redmond, Washington, USA) and either
adhered to the hair on the seal or attached to the rear flipper inter-digital webbing. The
use of bio-loggers adhered to the back or head provide some benefits over flipper mounted
devices (e.g. increased satellite transmittal rates, locations at sea) but these are lost
during the annual molt and limits the effective length of haul-out timelines.
Additionally, sensors attached to the head or dorsal region are often dry while the seal
is at the surface and can slightly bias the percent-dry values recorded by the device. For
this study, in cases where both tag types were deployed, percent dry observations from the
flipper tag were preferred. Sex and age class were determined at the time of
deployment by experts in the field. Seals determined to be less than one year were
classified as young of the year. subadults are those seals likely greater than one year of
age but not yet sexually mature. Adults are likely sexually mature and older than
approximately four years. For those tags deployed on young of the year and transmitting
into the next year, the age class was advanced to subadult on March 1 of the following
year. Table \@ref(tab:tblDeploy) provides a summary of these deployments and data received
from them.

@@

Tags that fall off due to molt or mortality and remain on ice or land can still send data
to satellites and we did not want to include data from these tags in our analyses. As
such, start and end times of each deployment were made by experts who examined several
simultaneous data streams (e.g., tag locations and dive behavior) to determine when tags
ceased providing data consistent with seal behavior. Sensor readings outside of these
start and end times were discarded prior to analysis. After approximately 9 months, a few
(7) devices deployed on the rear flipper of bearded seals reported implausible sensor
data. All data after the first instance of unrealistic values were censored from this
analysis. Figure \@ref(fig:dataCal) shows a calendar distribution of these data across the
study season for each species. Observations are concentrated in the months of May and June
due to the timing of deployment (April and May) and the timing of molt (May and June).
During molt, seals (and their attached bio-loggers) spend more time out of the water and
more data is transmitted. Molt timing also impacts when many deployments end as any tags
adhered to the hair will fall off.

@@

Haul-out behavior data was recorded and transmitted via the Argos satellite network as
hourly percent-dry timelines. For each hour of a day, the wet/dry sensor was polled by the
tag firmware every few seconds and a percent of the hour in a dry state was calculated
(\@ref(fig:examplePlot)). Each data transmission message consisted of a complete 24-hour
record. Tags were deployed and programmed in a manner to maximize data transmission during
the spring pupping and molting period. However, not all days during the deployment were
successfully transmitted. This is due to a variety of factors including satellite
coverage, tag availability (i.e. tags mounted to the rear flipper are not available to the
surface while the seal is at sea), tag performance, duty cycling, and extra-terrestrial
interference.

@@

In addition to sex and age class (non-dependent *young of the year*, sexually immature
*subadults*, and mature *adults*), we gathered additional variables that might help
explain variation in haul-out probabilities. These included day-of-year (for seasonal
effects) and solar hour (for diurnal effects). Solar hour was calculated using the {solaR}
package [@perpinan2012jss] within the R statistical environment [@rcoreteam2019] using the
daily locations. We also compiled several meteorological variables shown to affect
haul-out behavior in other Arctic pinnipeds [@reder2003pb; @udevitz2009cjz; @perry2017]. In
particular, we linked locations from satellite tags to key weather values from the North
American Regional Reanalysis (NARR) model produced by the National Centers for
Environmental Prediction [@mesinger2006bams]. The NARR model assimilates observational
data to produce a long-term picture of weather over North America. Numerous weather
variables are made available across the region 8 times daily. For this study, NARR weather
values were subset to the extent of our study area over the Bering and Chukchi Seas at
3-hr intervals based on a grid resolution of 32 km (1024 sq. km) The following
meteorological variables were interpolated and assigned to seal locations using a bilinear
method: 1) air temperature at 2m above the earth's surface, 2) wind consisting of
northerly and easterly vector components converted to wind speed using the Euclidean norm,
3) barometric pressure at sea level, and 4) precipitation (Table
\@ref(tab:tblCovariates)).

@@

We did explore whether the annual variation in maximum spring sea ice extent in the Bering
Sea influences the seasonal peak of seal haul-out behavior. In particular, we used sea ice
concentration data from the Nimbus-7 SMMR and DMSP SSM/I-SSMIS Passive Microwave Dataset,
Version 1 [@cavalieri1996] to calculate maximum sea ice extent. All sea ice concentration
grid cells (25 km^2^) in the study area with greater than 15% concentration were summed to
get total sea ice extent for each day between 15 February and 15 July across all years.
Maximum spring sea ice extent was simply the largest daily value for each year.

@@

To calculate plausible values for $\omega_{s,a}$ we obtained natural mortality predictions
from a hierarchical meta-analysis of phocid natural mortality [@trukhanova2018ee]. We then
assumed a 50/50 sex ratio at birth, 1 pup per litter, and reproductive schedules as
reported by Fedoseev [-@fedoseev2000] for seals in the western Bering Sea to formulate
Leslie matrices Ai for each species i. Stable age distributions
$v_i={v\_i0,v\_i1,⋯,v\_iA}$ were then determined by normalizing the dominant eigenvector
of $A_i$ so that they summed to one . Given a stable age distribution, we calculated the
proportion of the population by sex and stage (young of the year: 0-1 year; subadult: at
least one year but sexually immature; adult: sexually mature) using data on sexual
maturity from Fedoseev [-@fedoseev2000]. In particular, the expected proportion of the
population that are sex $s$ and young-of-year was set as $\pi_{i,s,yoy}=0.5v_i0$, while
the expected proportion that are subadults and adults were set to

@@

Ribbon seals show a pattern of gradually increasing haul-out probability in April that
peaks in May for subadults and in June for adults. The behavior is clearly centered around
local solar noon and expands to other hours later in the season as seals enter their
molting period. Subadults showed an earlier start and more intense haul-out activity in
April and May. The young of the year records begin after weaning and the model predictions
demonstrate the ontogeny of in-water activities (e.g. diving, foraging) in May. Adult
females have a more protracted season compared to males and more time spent resting in
June and July. The model suggests adult male ribbon seals complete their molt by the end
of June. Lastly, for adults of both sexes there is some indication for a shift to a
nocturnal haul-out pattern in late June and July.

@@

Compared to ribbon seals, spotted seals showed a longer spring haul-out season that is
less intensely centered on solar noon. Adults of both sexes spend considerable time in
April and May resting on the ice. As with ribbon seals, the young of the year records
begin after weaning and the model predictions demonstrate the ontogeny of in-water
activities (e.g. diving, foraging) in May.

@@

*Josh M. London:*

*Paul B. Conn:*

*Stacie K. Hardy:*

*Erin L. Richmond:*

*Jay M. Ver Hoef:*

*Irina S. Trukhanova:*

*Michael F. Cameron:*

*Peter L. Boveng:*

>>>>

Ice-associated seals rely on sea ice for a variety of activities, including breeding,
molting, pupping, and resting. In the Arctic, many of these activities occur in spring and
early summer (April-June) as sea ice begins to melt and retreat northward. Rapid
acceleration of climate change in Arctic ecosystems is therefore of concern as the
quantity and quality of suitable habitat is forecast to decrease. In this paper, we use
data collected from satellite-linked bio-loggers deployed between 2005 and 2020 to
investigate the seasonal timing and environmental factors affecting sea ice use by seals
(specifically, bearded, ribbon, and spotted seals) in the Bering and Chukchi seas. In
addition to providing baseline data on phenology, these data also allow us to refine
availability estimates needed to accurately estimate abundance from aerial survey counts
of seals basking on ice (i.e., to correct for the proportion of animals that are in the
water while surveys are conducted). Using generalized linear mixed pseudo-models to
properly account for temporal autocorrelation, we fit models with a variety of covariates
(e.g., day-of-year, solar hour, age-sex class, wind speed, barometric pressure,
temperature, precipitation) to examine their ability to explain variation in hourly
haul-out records. We found evidence for strong diurnal and within season patterns in
haul-out behavior, as well as strong weather effects (particularly wind and temperature).
In general, seals were more likely to rest on ice in the middle of the day and when wind
speed was low and temperatures were higher. Haul-out propensity increased through March
and April, peaking in May and early June before declining again. The timing and frequency
of haul-out events also varied based on species and age-sex class. For ribbon and spotted
seals, models with year effects were highly supported, indicating that the timing and
magnitude of haul-out behavior varied among years. However, we did not find evidence that
haul-out timing was linked annual sea ice extent. Our analysis emphasizes the importance
of accounting for variation in haul-out behavior when interpreting the number of seals
counted in aerial surveys.

@@

For ice-associated seals, haul out is tightly linked with sea ice. Ribbon seals rest out
of the water almost exclusively on sea ice and are mostly pelagic outside the spring
pupping, breeding, and molting season [@burns1981]. While spotted and bearded seals do
rest on beaches during ice-free months, they strongly prefer sea ice as a resting platform
during the spring and early summer [@burns2009]. The very remote nature of the Bering and
Chukchi seas mean direct observation of seal behavior is limited. Bio-logging devices are,
thus, especially useful tools for collecting key information on movement and haul-out
behavior for these species. Records of time spent out of the water provide valuable data
for identifying covariates that explain variation in haul-out behavior. For instance, in
the Antarctic, Bengtson and Cameron [-@bengtson2004pb] relied on bio-logging data to
demonstrate greater haul-out propensity in juvenile crabeater seals (*Lobodon
carcinophaga*) than adults, with highest probabilities in February and at times close to
solar noon. In the Arctic, Von Duyke et al. [-@vonduyke2020ee] demonstrated seasonal
changes between diurnal and nocturnal haul-out behavior of ringed seals. Bengtson et al.
[-@bengtson2005pb] documented a higher propensity for ringed seal basking near solar noon,
as did Ver Hoef et al. [-@verhoef2014sm] in an analysis of bearded (*Erignathus
barbatus*), ribbon (*Histriophoca fasciata*), and spotted (*Phoca largha*) seals using
much larger sample sizes. Olnes et al. [-@olnes2020pb] showed that the proportion of time
bearded seals spent hauled out progressively increased through spring and summer. And, Ver
Hoef et al. [-@verhoef2014sm] found haul-out probabilities increased gradually starting in
March and peaking in May and June for bearded, ribbon, and spotted seals. However, their
primary focus was to develop a correction factor for estimating seal abundance, and like a
subsequent analysis by Conn et al. [-@conn2014mee], they did not attempt to separately
analyze haul-out schedules for different age and sex classes (although initial models with
sex and age as fixed effects were insignificant at p = 0.05; [@verhoef2014sm]).

@@

Ultimately, knowledge of trends in phenology and abundance (or life history surrogates
such as survival and recruitment) will be necessary to make credible quantitative
predictions about the effects of the climate crisis on the abundance and distribution of
Arctic seal populations. Before we can construct a trend, however, we first require a
baseline. Several studies have contributed estimates of the distribution and abundance of
ice-associated seal species in the Arctic using aerial surveys (e.g., [@bengtson2005pb],
[@conn2014mee], and [@verhoef2014sm]). Such studies are conducted over very large areas
and estimation of absolute abundance requires making inference about numerous nuisance
processes affecting the observation of seals on ice. This includes availability (only
seals basking on ice are available to be counted), detection probability (observers or
automated detection systems may miss some seals on ice), species misclassification, and
possible disturbance of seals by aircraft [@conn2014mee;
@verhoef2014sm].

@@

Haul-out behavior data from `r dplyr::n_distinct(deploy_tbl$speno)` bio-loggers deployed
on bearded, spotted, and ribbon seals was subset to include only records from 1 March to
15 July between `r min(lubridate::year(dat.sf$haulout_dt))` and
`r max(lubridate::year(dat.sf$haulout_dt))`. Bio-loggers were of the 'SPLASH' or 'SPOT'
family of tags developed by Wildlife Computers (Redmond, Washington, USA) and either
adhered to the hair on the seal or attached to the rear flipper inter-digital webbing. The
use of bio-loggers adhered to the back or head provide some benefits over flipper mounted
devices (e.g. increased satellite transmittal rates, locations at sea) but these are lost
during the annual molt and limits the effective length of haul-out timelines.
Additionally, sensors attached to the head or dorsal region are often dry while the seal
is at the surface and can slightly bias the percent-dry values recorded by the device. For
this study, in cases where both tag types were deployed, percent dry observations from the
flipper tag were preferred. Sex and age class were determined at the time of
deployment by experts in the field. Seals determined to be less than one year were
classified as young-of-the-year. subadults are those seals likely greater than one year of
age but not yet sexually mature. Adults are likely sexually mature and older than
approximately four years. For those tags deployed on young-of-the-year and transmitting
into the next year, the age class was advanced to subadult on March 1 of the following
year. Table \@ref(tab:tblDeploy) provides a summary of these deployments and data received
from them.

@@

Tags that fall off due to molt or mortality and remain on ice or land can still send data
to satellites and we did not want to include data from these tags in our analyses. As
such, start and end times of each deployment were identified by experts who examined several
simultaneous data streams (e.g., tag locations and dive behavior) to determine when tags
ceased providing data consistent with seal behavior. Sensor readings outside of these
start and end times were discarded prior to analysis. After approximately 9 months, a few
(7) devices deployed on the rear flipper of bearded seals reported implausible sensor
data. All data after the first instance of unrealistic values were censored from this
analysis. Figure \@ref(fig:dataCal) shows a calendar distribution of these data across the
study season for each species. Observations are concentrated in the months of May and June
due to the timing of deployment (April and May) and the timing of molt (May and June).
During molt, seals (and their attached bio-loggers) spend more time out of the water and
more data is transmitted. Molt timing also impacts when many deployments end as any tags
adhered to the hair will fall off.

@@

Haul-out behavior data was recorded and transmitted via the Argos satellite network as
hourly percent-dry timelines. For each hour of a day, the wet/dry sensor was polled by the
tag firmware every few seconds and a percent of the hour in a dry state was calculated
(Figure \@ref(fig:examplePlot)). Each data transmission message consisted of a complete 24-hour
record. Tags were deployed and programmed in a manner to maximize data transmission during
the spring pupping and molting period. However, not all days during the deployment were
successfully transmitted. This is due to a variety of factors including satellite
coverage, tag availability (i.e. tags mounted to the rear flipper are not available to the
surface while the seal is at sea), tag performance, duty cycling, and extra-terrestrial
interference.

@@

In addition to sex and age class (non-dependent *young-of-the-year*, sexually immature
*subadults*, and mature *adults*), we gathered additional variables that might help
explain variation in haul-out probabilities. These included day-of-year (for seasonal
effects) and solar hour (for diurnal effects). Solar hour was calculated using the {solaR}
package [@perpinan2012jss] within the R statistical environment [@rcoreteam2019] using the
daily locations. We also compiled several meteorological variables shown to affect
haul-out behavior in other Arctic pinnipeds [@reder2003pb; @udevitz2009cjz; @perry2017]. In
particular, we linked locations from satellite tags to key weather values from the North
American Regional Reanalysis (NARR) model produced by the National Centers for
Environmental Prediction [@mesinger2006bams]. The NARR model assimilates observational
data to produce a long-term picture of weather over North America. Numerous weather
variables are made available across the region 8 times daily. For this study, NARR weather
values were subset to the extent of our study area over the Bering and Chukchi Seas at
3-hr intervals based on the native grid resolution of 32 km (1024 sq. km). The following
meteorological variables were interpolated and assigned to seal locations using a bilinear
method: 1) air temperature at 2m above the earth's surface, 2) wind consisting of
northerly and easterly vector components converted to wind speed using the Euclidean norm,
3) barometric pressure at sea level, and 4) precipitation (Table
\@ref(tab:tblCovariates)).

@@

We did explore whether the annual variation in maximum spring sea ice extent in the Bering
Sea influences the seasonal peak of seal haul-out behavior. In particular, we used sea ice
concentration data from the Nimbus-7 SMMR and DMSP SSM/I-SSMIS Passive Microwave Dataset,
Version 1 [@cavalieri1996] to calculate maximum sea ice extent. All sea ice concentration
grid cells (25 km^2^) in the study area with greater than 15% concentration were summed to
get the total sea ice extent for each day between 15 February and 15 July across all years.
Maximum spring sea ice extent was simply the largest daily value for each year.

@@

To calculate plausible values for $\omega_{s,a}$ we obtained natural mortality predictions
from a hierarchical meta-analysis of phocid natural mortality [@trukhanova2018ee]. We then
assumed a 50/50 sex ratio at birth, 1 pup per litter, and reproductive schedules as
reported by Fedoseev [-@fedoseev2000] for seals in the western Bering Sea to formulate
Leslie matrices Ai for each species i. Stable age distributions
$v_i={v\_i0,v\_i1,⋯,v\_iA}$ were then determined by normalizing the dominant eigenvector
of $A_i$ so that they summed to one . Given a stable age distribution, we calculated the
proportion of the population by sex and stage (young-of-the-year: 0-1 year; subadult: at
least one year but sexually immature; adult: sexually mature) using data on sexual
maturity from Fedoseev [-@fedoseev2000]. In particular, the expected proportion of the
population that are sex $s$ and young-of-year was set as $\pi_{i,s,yoy}=0.5v_i0$, while
the expected proportion that are subadults and adults were set to

@@

Ribbon seals show a pattern of gradually increasing haul-out probability in April that
peaks in May for subadults and in June for adults. The behavior is clearly centered around
local solar noon and expands to other hours later in the season as seals enter their
molting period. Subadults showed an earlier start and more intense haul-out activity in
April and May. The young-of-the-year records begin after weaning and the model predictions
demonstrate the ontogeny of in-water activities (e.g. diving, foraging) in May. Adult
females have a more protracted season compared to males and more time spent resting in
June and July. The model suggests adult male ribbon seals complete their molt by the end
of June. Lastly, for adults of both sexes there is some indication for a shift to a
nocturnal haul-out pattern in late June and July.

@@

Compared to ribbon seals, spotted seals showed a longer spring haul-out season that is
less intensely centered on solar noon. Adults of both sexes spend considerable time in
April and May resting on the ice. As with ribbon seals, the young-of-the-year records
begin after weaning and the model predictions demonstrate the ontogeny of in-water
activities (e.g. diving, foraging) in May.

@@

*Josh M. London:* led field work, conceptualization, methodology, formal analysis, code development, writing of original draft, visualization, review, revisions, and corresponding author

*Paul B. Conn:* conceptualization, methodology, formal analysis, code development, writing of original draft, review, and revisions

*Stacie K. Hardy:* field work, methodology, data curation, validation, review, and revisions

*Erin L. Richmond:* data curation, conceptualization, methodology, review, and revisions

*Jay M. Ver Hoef:* field work, conceptualization, methodology, code development, review, and revisions

*Irina S. Trukhanova:* conceptualization, formal analysis, code development, review, and revisions

*Michael F. Cameron:* led field work, project administration, methodology, review, and revisions

*Peter L. Boveng:* led field work, supervision, project administration, conceptualization, methodology, review, and revisions
